\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2022-06-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
address comments from reviewers

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of independent censoring is treated in Saarela [-@saarela2016case]). We assume a continuous time model, which implies that the counting process jumps are less than or equal to one. We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.

@@

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
Therefore, an estimating function for $\theta$ can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\theta)$.
  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\theta)\right)$.
  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.

@@

plotb <- emptyPlot + geom_point(data = dplyr::filter(a, event == 1),
                                aes(x = time, y = ycoord, 
                                    color = "Case series",
                                    fill = "Case series"), 
                                show.legend = FALSE, size = 1.5, 
                                alpha = 0.5, shape = 21) + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = label_number_si())

plotc <- plot(a, add.case.series = TRUE, legend = FALSE) +
  scale_y_continuous(labels = label_number_si())

@@

legend_b <- get_legend(
  plotd + 
    guides(color = guide_legend(nrow = 1)) +
     theme(legend.direction = "horizontal",
           legend.justification = "center",
           legend.box.just = "bottom")
)

@@

From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals. These quantities appear in Table \ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is 48% lower than that for ALL. Moreover, being in relapse at transplant is associated with a hazard ratio of 4.09 when compared to CR1.

>>>>

To describe the theoretical foundations of case-base sampling, we use the framework of counting processes. In what follows, we abuse notation slightly and omit any mention of $\sigma$-algebras. Instead, following Aalen \textit{et al} [-@aalen2008survival], we use the placeholder "past" to denote the past history of the corresponding process. The reader interested in more details can refer to Saarela & Arjas [-@saarela2015non] and Saarela [-@saarela2016case]. First, let $N_{i}(t) \in \{0, 1\}$ be counting processes corresponding to the event of interest for individual $i=1, \ldots,n$. For simplicity, we will consider Type I censoring due to the end of follow-up at time $\tau$ (the general case of independent censoring is treated in Saarela [-@saarela2016case]). We are interested in modeling the hazard functions $\lambda_{i}(t)$ of the processes $N_i(t)$, and which satisfy
$$\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].$$  The processes $N_i(t)$ count the person-moments from the *case series*.

@@

Case-base sampling allows us to avoid this integral. By conditioning on a sampled person-moment, we get individual partial likelihood contributions of the form
$$P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.$$
By using the symbol $\stackrel{\theta}{\propto}$, we mean that both sides of the expression are equal up to multiplicative factors that do not depend on $\theta$. Therefore, an estimating function for $\theta$ can be composed of these contributions as:
\begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation}
When a logarithmic link function is used for modeling the hazard function, the above expression is of a logistic regression form with an offset term $\log(1/\rho_i(t))$. Note that the sampling units selected in the case-base sampling mechanism are person-moments, rather than individuals, and the parameters to be estimated are hazards or hazard ratios rather than odds or odds ratios. Generally, an individual can contribute more than one person-moment, and thus the terms in the product integral are not independent. Nonetheless, Saarela [-@saarela2016case] showed that the corresponding partial likelihood score function has mean zero at the true value $\theta=\theta_0$, and that the resulting estimator $\hat{\theta}$ is asymptotically normally distributed.

@@

  1. Compute $\hat{\theta}$ and its variance-covariance matrix $V(\hat \theta)$ which is obtained as part of the logistic regression output.
  2. Sample $B$ times from a multivariate normal $N\left(\hat{\theta}, V(\hat\theta)\right)$.
  3. For each sample, compute the survival function using Equation \ref{eqn:surv} below.
  4. Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.

@@

plotb <- emptyPlot + geom_point(data = dplyr::filter(a, event == 1),
                                aes(x = time, y = ycoord, 
                                    color = "Case series",
                                    fill = "Case series"), 
                                show.legend = FALSE, size = 1.5, 
                                alpha = 0.5, shape = 21) + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = label_number_si()) +
  scale_color_manual(values = c("#AB3A59"),
                     limits = c("Case series"),
                     name = "") +
  scale_fill_manual(values = c("#E16A86"),
                    limits = c("Case series"),
                    name = "")

plotc <- plot(a, add.case.series = TRUE, legend = FALSE) +
  scale_y_continuous(labels = label_number_si()) +
  scale_color_manual(values = c("#AB3A59"),
                     limits = c("Case series"),
                     name = "") +
  scale_fill_manual(values = c("#E16A86"),
                    limits = c("Case series"),
                    name = "")

@@

legend_b <- get_legend(
  plotd + 
    theme(legend.direction = "horizontal",
           legend.justification = "center",
           legend.box.just = "bottom") +
    scale_color_manual(values = c("#AB3A59", "#347004"),
                       limits = c("Case series", "Base series"),
                       name = "") +
    scale_fill_manual(values = c("#E16A86", "#50A315"),
                      limits = c("Case series", "Base series"),
                      name = "") +
      guides(color = guide_legend(nrow = 1),
             fill = guide_legend(nrow = 1))
)

@@

From the fit object, we can extract both the hazard ratios and their corresponding confidence intervals. These quantities appear in Table \ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to a significant hazard ratio: the hazard for AML is about half that for ALL. Moreover, being in relapse at transplant is associated with a hazard ratio of 4.22 when compared to CR1.

### 2022-06-06 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex
final final
address comments from reviewers

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
independent censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
The processes \(N_i(t)\) count the person-moments from the \emph{case
series}.

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual partial likelihood
contributions of the form
\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, an estimating function for \(\theta\) can be composed of
these contributions as: \begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation} When a logarithmic link function is used for modeling the
hazard function, the above expression is of a logistic regression form
with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
selected in the case-base sampling mechanism are person-moments, rather
than individuals, and the parameters to be estimated are hazards or
hazard ratios rather than odds or odds ratios. Generally, an individual
can contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the corresponding partial
likelihood score function has mean zero at the true value
\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
is asymptotically normally distributed.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\hat{\theta}\) and its variance-covariance matrix
  \(V(\theta)\).
\item
  Sample \(B\) times from a multivariate normal
  \(N\left(\hat{\theta}, V(\theta)\right)\).
\item
  For each sample, compute the survival function using Equation
  \ref{eqn:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to
  construct a pointwise confidence band for the survival function of
  interest.
\end{enumerate}

@@

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
a significant hazard ratio: the hazard for AML is 48\% lower than that
for ALL. Moreover, being in relapse at transplant is associated with a
hazard ratio of 4.09 when compared to CR1.

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
\url{http://sahirbhatnagar.com/}%
%
\\\email{sahir.bhatnagar@mcgill.ca}
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}%
%
\\\email{max.turgeon@umanitoba.ca}
}

\address{%
Jesse Islam\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
%
%
\\\email{jesse.islam@mail.mcgill.ca}
}

\address{%
James A. Hanley\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}%
%
\\\email{james.hanley@mcgill.ca}
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}%
%
\\\email{olli.saarela@utoronto.ca}
}

>>>>

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \textit{et al} \citeyearpar{aalen2008survival}, we use the
placeholder ``past'' to denote the past history of the corresponding
process. The reader interested in more details can refer to Saarela \&
Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
independent censoring is treated in Saarela
\citeyearpar{saarela2016case}). We are interested in modeling the hazard
functions \(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which
satisfy
\[\lambda_{i}(t)\,\textrm dt = E[\,\textrm dN_{i}(t)\mid \mathrm{past}].\]
The processes \(N_i(t)\) count the person-moments from the \emph{case
series}.

@@

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual partial likelihood
contributions of the form
\[P(\,\textrm dN_{i}(t) \mid \,\textrm dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
By using the symbol \(\stackrel{\theta}{\propto}\), we mean that both
sides of the expression are equal up to multiplicative factors that do
not depend on \(\theta\). Therefore, an estimating function for
\(\theta\) can be composed of these contributions as: \begin{equation}
L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{\,\textrm dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{\,\textrm dQ_i(t)}. \label{eq:lik-function}
\end{equation} When a logarithmic link function is used for modeling the
hazard function, the above expression is of a logistic regression form
with an offset term \(\log(1/\rho_i(t))\). Note that the sampling units
selected in the case-base sampling mechanism are person-moments, rather
than individuals, and the parameters to be estimated are hazards or
hazard ratios rather than odds or odds ratios. Generally, an individual
can contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the corresponding partial
likelihood score function has mean zero at the true value
\(\theta=\theta_0\), and that the resulting estimator \(\hat{\theta}\)
is asymptotically normally distributed.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\hat{\theta}\) and its variance-covariance matrix
  \(V(\hat \theta)\) which is obtained as part of the logistic
  regression output.
\item
  Sample \(B\) times from a multivariate normal
  \(N\left(\hat{\theta}, V(\hat\theta)\right)\).
\item
  For each sample, compute the survival function using Equation
  \ref{eqn:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to
  construct a pointwise confidence band for the survival function of
  interest.
\end{enumerate}

@@

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the type of disease corresponds to
a significant hazard ratio: the hazard for AML is about half that for
ALL. Moreover, being in relapse at transplant is associated with a
hazard ratio of 4.22 when compared to CR1.

@@

\address{%
Sahir Rai Bhatnagar*\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
\url{http://sahirbhatnagar.com/}\\%
%
\email{sahir.bhatnagar@mcgill.ca}%
}

\address{%
Maxime Turgeon*\\
University of Manitoba\\%
186 Dysart Road Winnipeg, MB, Canada R3T 2N2\\
%
\url{https://maxturgeon.ca/}\\%
%
\email{max.turgeon@umanitoba.ca}%
}

\address{%
Jesse Islam\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
%
%
\email{jesse.islam@mail.mcgill.ca}%
}

\address{%
James A. Hanley\\
McGill University\\%
2001 McGill College Avenue Montreal, QC, Canada H3A 1G1\\
%
\url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\%
%
\email{james.hanley@mcgill.ca}%
}

\address{%
Olli Saarela\\
University of Toronto\\%
Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
%
\url{http://individual.utoronto.ca/osaarela/}\\%
%
\email{olli.saarela@utoronto.ca}%
}
